---
title: "RepData_PeerAssessement2"
author: "Kurt Fitz"
date: "9/3/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#Analysis of weather events resulting in greatest harm to American population health.

## Synopsis
TODO: Synopsis: Immediately after the title, there should be a synopsis which describes and summarizes your analysis in at most 10 complete sentences.

The following summary describes the analysis performed on the [NOAA (National Oceanic and Atmospheric Administration)](http://noaa.gov/)
) Storm Database dataset. The Analysis is focused on identifying the storm events resulting in the greatest impact to population health.

## Data Processing
TODO: There should be a section titled Data Processing which describes (in words and code) how the data were loaded into R and processed for analysis. In particular, your analysis must start from the raw CSV file containing the data. You cannot do any preprocessing outside the document. If preprocessing is time-consuming you may consider using the \color{red}{\verb|cache = TRUE|}cache=TRUE option for certain code chunks.

Download and load the dataset into memory.
```{r}
# Download the datset as zip file.
if(!file.exists("data/stormdata.zip"))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/stormdata.zip")
  
# Load the datset into memory 
dataset <- read.csv("data/stormdata.zip")
```

After loading the dataset into memory lets look at it's dimensions and classes.
```{r}
dims <- dim(dataset)

colClasses <- sapply(dataset, classes)
```
The dataset contains `r dims[1]` rows and `r dims[2]` columns. Next, we can look at the columns of the classes:
`r colClasses`

Questions
Your data analysis must address the following questions:

Across the United States, which types of events (as indicated in the \color{red}{\verb|EVTYPE|}EVTYPE variable) are most harmful with respect to population health?
Across the United States, which types of events have the greatest economic consequences?

## Results
TODO: There should be a section titled Results in which your results are presented.
You may have other sections in your analysis, but Data Processing and Results are required.

The analysis document must have at least one figure containing a plot.
Your analysis must have no more than three figures. Figures may have multiple plots in them (i.e. panel plots), but there cannot be more than three figures total.
You must show all your code for the work in your analysis document. This may make the document a bit verbose, but that is okay. In general, you should ensure that \color{red}{\verb|echo = TRUE|}echo=TRUE for every code chunk (this is the default setting in knitr).

Assignment
The basic goal of this assignment is to explore the NOAA Storm Database and answer some basic questions about severe weather events. You must use the database to answer the questions below and show the code for your entire analysis. Your analysis can consist of tables, figures, or other summaries. You may use any R package you want to support your analysis.



