---
title: "Analysis of weather events resulting in greatest impact to population health and economic factors in America."
author: "Kurt Fitz"
date: "10/01/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(ggplot2)
library(grid)
library(gridExtra)
```


The following summary describes the analysis performed on the [NOAA (National Oceanic and Atmospheric Administration)](http://noaa.gov/)
) Storm Database dataset. The Analysis is focused on identifying the storm events resulting in the greatest cost to the economy and population health.

## Data Processing

Download and load the dataset into memory.
```{r cache=TRUE }
# Download the datset as zip file.
if(!file.exists("data/stormdata.zip"))
  download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "data/stormdata.zip")
  
# Load the datset into memory 
dataset <- read.csv("data/stormdata.zip")
```

After loading the dataset into memory lets look at it's dimensions and classes.
```{r}
dims <- dim(dataset)

colClasses <- sapply(dataset, class)
```

The dataset contains `r dims[1]` rows and `r dims[2]` columns. Next, we can look at the columns of the classes:
`r colClasses`.

Next, to determine which events are most harmful with respect to population health we can examine the dataset to determine which events resulted in the greatest number of fatalities and injuries.
```{r}
#subset the data to fatalities, injuries, types of economic damage, and event type.
damageData <- dataset %>% select(EVTYPE, FATALITIES, INJURIES, CROPDMG, CROPDMGEXP, PROPDMG, PROPDMGEXP)

```

### Convert PROPDMGEXP and CROPDMGEXP variables to dolllar amounts
The PROPDMGEXP and CROPDMGEXP columns serve as multipliers to the CROPDMG and PROPDMG columns and we need to process them to their dollar values to make easier to understand dollar value comparisons.
```{r}
EXP.level <- c("", "-", "?", "+", 0:9, "B", "b", "H", "h", "K", "k", "M", "m")
EXP.level.fixed <- c(0, 0, 0, 0, 0:9, 9, 9, 2, 2, 3, 3, 6, 6)
damageData$CROPDMGEXP <-  plyr::mapvalues(damageData$CROPDMGEXP, from = EXP.level, to = EXP.level.fixed, warn_missing = F)
damageData$PROPDMGEXP <-  plyr::mapvalues(damageData$PROPDMGEXP, from = EXP.level, to = EXP.level.fixed, warn_missing = F)

damageData$CROPDMGEXP <- as.numeric(damageData$CROPDMGEXP)
damageData$PROPDMGEXP <- as.numeric(damageData$PROPDMGEXP)

damageData$PROPDMG = damageData$PROPDMG * 10 ^ damageData$PROPDMGEXP
damageData$CROPDMG = damageData$CROPDMG * 10 ^ damageData$CROPDMGEXP

damageData <- subset(damageData, select = -c(PROPDMGEXP, CROPDMGEXP))
```


Following is a summary of the dataset containing the fatalities and injuries by event type.
```{r}
summary(damageData)
```


## Results
### Weather Events and their effects on Health
Next in our anaylsis, we create a plot of the weather events that result in greatest number of fatalities and injuries.
```{r}

damageData <- group_by(damageData, EVTYPE)

injuries <- damageData %>% mutate(TOTAL_INJURIES=sum(INJURIES)) %>% filter(TOTAL_INJURIES > 500)

injuriesPlot <- ggplot(injuries, aes(x=EVTYPE, y=INJURIES))+ 
geom_bar(stat='identity',position="dodge")+
labs(x="Weather Event",y="Total Injuries")+
theme(axis.text.x=element_text(angle=45, hjust=1))


fatalities <- damageData %>% mutate(TOTAL_FATALITIES=sum(FATALITIES))

fatalitiesPlot <- ggplot(fatalities %>% filter(TOTAL_FATALITIES > 200), aes(x=EVTYPE, y=FATALITIES))+ 
geom_bar(stat='identity',position="dodge")+
labs(x="Weather Event",y="Total Fatalities")+
theme(axis.text.x=element_text(angle=45, hjust=1))

plots <- grid.arrange(fatalitiesPlot, injuriesPlot, nrow=1)
#print(plots)

```

From the resulting plot, we can see that Tornados and Heat realated weather events resulted in the greatest number of fatalities, while Ice Storms and Tornados resulted in the greatest number of injuries.

### Weather Events and their Economic Consequences

```{r}
propDamage <- damageData %>% mutate(PROP_DMG = sum(PROPDMG)) %>% filter(PROP_DMG > 100000000000)

propDamagePlot <- ggplot(propDamage, aes(x=EVTYPE, y=PROPDMG))+ 
geom_bar(stat='identity',position="dodge")+
labs(x="Weather Event",y="Property Damage")+
theme(axis.text.x=element_text(angle=45, hjust=1))

cropDamage <- damageData %>% mutate(CROP_DMG = sum(CROPDMG)) %>% filter(CROP_DMG > 100000000)

cropDamagePlot <- ggplot(cropDamage, aes(x=EVTYPE, y=CROPDMG))+ 
geom_bar(stat='identity',position="dodge")+
labs(x="Weather Event",y="Crop Damage")+
theme(axis.text.x=element_text(angle=45, hjust=1))

plots <- grid.arrange(propDamagePlot, cropDamagePlot, nrow=1)

```

From the resulting plot, we can see that floods and storm surges are the most detrimental with regard to property damage, while drought and extreme cold have the greatest negative impact on crops.


